Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-21 text/manuscript-full-KS.qmd
went through results

### Decentralized Diversity and Income in a Multi-Nucleo Landscape

[START HERE] Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles -- 11 Geomarkets, results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), which are shown in the Online Appendix. The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core (CITE- Harris and Ullman 1945). As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

The population growth in the Los Angeles metropolitan area increased from approximately 8 Million in 1980 to more than 12 Million in 2020. @fig-la-maps showcases population growth by Geomarkets over the three census periods reflect the urban sprawl forming the multi-nuclei model. The most pronounced increases occurred in South and South Central Los Angeles Geomarket, as well as the outer/inland Geomarkets of San Fernando Valley-West, Riverside and San Bernadino, and Ontario. The South and South Central Los Angeles Geomarket grew from approximately 1.2 to 1.6 Million from 1980 to 2020. In contrast, the San Fernando Valley-West and Riverside and San Bernadino, and Ontario Geomarkets grew from approximately 700,000 to 1.2 Million and from 1.1 Million to nearly 3 Million during this time, respectively.

@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. Over this period, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. \[THE CAVEAT THAT WE DON'T HAVE ASIAN FOR 1980 IS MORE IMPORTANT FOR LA\] In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 62% White in Long Beach to 78% in the San Fernando Valleys (West and East) Geomarkets. By 2020, however, only the West Los Angeles and West Beach Geomarket remained predominantly White (62%). Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (44% White, 4% Black, 39% Hispanic; 11% Asian Pacific Islander), San Fernando Valley -East (43% White, 4% Black, 41% Hispanic; 8% Asian Pacific Islander), Glendale and Pasadena (35% White, 3% Black, 37% Hispanic; 21% Asian Pacific Islander), South Bay (29% White, 12% Black, 33% Hispanic; 21% Asian Pacific Islander), Long Beach (26% White, 11% Black, 46% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 3% Black, 52% Hispanic; 24% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 54% Hispanic; 6% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 18% and 18% to 9%, respectively.

@@

@fig-la-rq1-ses also illustrates parallel trends in poverty and educational attainment over time, with Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles Geomarkets consistenly ranking at lower levels across these socioeconomic indicators. Moreover, the disparities between these three Geomarkets and all others in the Los Angeles metropolitan area widened from 1980 to 2020.

<!--# In 1980, poverty rates were relatively low (less than 12%) for all Geomarkets except for East Los Angeles (17%), South and South Central Los Angeles (24%), and Hollywood and Wilshire (18%). However, poverty rates declined in South and South Central Los Angeles (22%), remained stable in East Los Angeles and Hollywood and Wilshire, and increased in all other Geomarkets. Such shifts and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in more staggered poverty rates across the metropolitan area by 2020. This includes four Geomarkets with poverty rates less than 10% (San Fernando Valley- West, West Los Angeles and West Beach, South Bay, and Covina and West Covina), four Geomarkets with poverty rates ranging from 11%-15% (San Fernando Valley- East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario), and three Geomarkets with poverty rates ranging from 16%-22% (Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles). -->

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets in the Los Angeles metropolitan areas, we analyze six orders placed by a research university that filtered for prospects across the entire state of California, in the 2020-2022 high school graduating classes, and by PSAT scores. Two orders indicated PSAT score thresholds ranging from 1070-1180, another two orders filtered for a 1190-1260 PSAT range, and the remaining two orders filtered for scores ranging from 1270-1520.

@fig-la-rq2-race presents the racial/ethnic composition of purchased student profiles. For the more than 15,000 prospects in the resulting student lists at the low PSAT score thresholds (1070-1180), South and South Central Los Angeles (n=1,031) and East Los Angeles (n=871) Geomarket pools stand out as having an overwhelming majority of students of color, 98% and 95% respectively. Relative to other Geomarkets in the Los Angeles metropolitan area, such patterns are attributed to both South and South Central Los Angeles (89%) and East Los Angeles (64%) having substantially greater proportions of Hispanic prosepcts, as well as a larger proportion of Black prospects for South and South Central Los Angeles (6%) and larger share of Asian prospects for East Los Angeles (30%).

Even among higher PSAT score ranges, South and South Central Los Angeles remains the most prominent Geomarket with significant proportions of Hispanic and Black student prospects, while East Los Angeles stands out for its substantial share of Asian prospects.

@@

@fig-la-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions for the Los Angeles metropolitan area, demonstrating South and South Central Los Angeles (7% versus 2% for all prospects) and Riverside, San Bernadino, and Ontario (25% versus 17% for all prospects) Geomarkets contributed larger proportions of Black prospects relative to their contributions to the overall prospect pool in the Los Angeles metropolitan area. <!--#  For the more than 2,500 White prospects in this pool, we see the San Fernando Valley - West (28% versus 20% for all prospects) and the West Los Angeles and West Beach (12% versus 7% for all prospects)Geomarkets contributing disproportionately larger shares.-->

For Hispanic prospects, South and South Central Los Angeles (8% versus 2% for all prospects) and Riverside, San Bernadino, and Ontario (24% versus 17% for all prospects) also contributed a disproportinately larger proportions of prospects relative to their contributions to the overall pool in the Los Angeles metropolitan area.

<!--# [I THINK CLEANER IF YOU JUST HAVE A SMALL PARAGRAPH FOR HISPANIC AND A SMALL PARAGRAPH FOR BLACK. I THINK ONE LEDE ACROSS THESE RESULTS IS RIVERSIDE AS AN ENGINE FOR BLACK/LATINX AND FOR FIRST-GEN STUDENTS. I GET THAT THESE TWO POINTS CONTRADICT EACH OTHER]. Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-la-rq2-midSAT-race. South and South Central Los Angeles and Riverside, San Bernadino, and Ontario Geomarkets contributed larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Los Angeles metropolitan area. On the other hand, the San Fernando Valley- West and the Glendale and Pasadena Geomarket contributed smaller proportions of Black and Hispanic Prospects. However, the West Los Angeles and West Beach (10% versus 7% for all prospects), Hollywood Wilshire (7% versus 5% for all prospects), and the Long Beach (8% versus 5% for all prospects) Geomarkets contributed a larger share of Black prospects, whereas Covina and West Covina contributed a larger share of Hispanic students. All other Geomarkets contributed nearly equal shares of Black and Hispanic prospects. -->

In comparison, the Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger share of Asian prospects relative to their contributions to the overall prospect pool.

These contribution patterns by race/ethnicity demonstrate that student list orders targeting prospects in the Los Angeles metropolitan area using Geomarket filters that exclude lower-income Geomarkets such as South and South Central Los Angeles and Riverside, San Bernadino, and Ontario Geomarkets would result in the loss of substantially disproportionate shares of Black and Hispanic students from the resulting prospect pools. Similarly, using a Geomarket filter excluding East Los Angeles would result in the loss of disproportionate shares of Asian students.

<!--# [I'M NOT SURE WHERE YOU ARE GETTING THESE NUMBERS FROM; THIS IS THE WRONG FIGURE I THINK. AND WHEN YOU REVISE, THE LEDES ARE RIVSERSIDE, EAST LA AND, SOUTH CENTRAL (ONLY 2% OF PROSPECTS BUT 10% OF NO-COLLEGE) KS-- weird... overall substantive rankings are correct but my %s are all off]. --> @fig-la-rq2-midSAT-ses presents Geomarket contributions by first-generation college student status, demonstrating the disproportionately larger share of first-generation college students contributed to the overall pool by East Los Angeles, South and South Central Los Angeles, and Riverside, San Bernadino, and Ontario Geomarkets. For students whose parents did not attend college, East Los Angeles (16% versus 5% of all prospects), South and South Central Los Angeles (10% versus 2% of all prospects), and Riverside, San Bernardino, and Ontario (24% versus 17% of all prospects) Geomarkets provided a disproportionately larger contribution of first-generation college students. These three Geomarkets also made disproportionately larger contributions of first-generation college student prospects whose parents attended but did not complete their degree, as well as Hollywood & Wilshire (7% versus 5% of all prospects), Long Beach (7% versus 5% of all prospects), and Covina and West Covina (13% versus 10% of all prospects).

<!--#  [I THINK THE WAY YOU WRITE UP THIS FIGURE SHOULD BE CHANGED MORE SUBSTANTIALLY; I ADD SOME COMMENTS BELOW] --> Lastly, @fig-la-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Los Angeles metropolitan area.

For Hispanic prospects, the South and South Central Los Angeles Geomarket stands out as providing the most substantial share of first-generation college students. Across all Hispanic prospects in the middle range PSAT orders, 25% were first-generation college students whose parents did not attend college, 25% were first-generation college students whose parents attended some college but did not complete their degrees, and 50% were not first-generation college students. In comparison, the South and South Central Los Angeles Geomarket contributed much larger disproportionate share of Hispanic first-generation college students whose parents did not attend college (54%), with the Hollywood & Wilshire (30%), East Los Angeles (39%), and Riverside, San Bernardino, and Ontario (31%) also contributing larger but more modest disproportionate shares. The East Los Angeles (33%), South and South Central Los Angeles (28%), Long Beach (30%), and Riverside, San Bernardino, and Ontario (27%) also contributed larger Hispanic first-generation college students whose parents attended college but did not complete their their degree.

@@

While the pool of Black prospects is relatively small (n=186) in @fig-la-rq2-midSAT-combo , some patterns emerge by first-generation status. Overall, approximately 5% of Black prospects were first-generation college students-no college, 20% were first-generation college students-some college, and 74% not first-generation). The South and South Central Los Angeles Geomarket stands out for contributing a disproportionately high share (21%) of Black first-generation college students whose parents did not attend college, despite representing a relatively smaller overall number. Additionally, the Long Beach Geomarket is notable for its higher proportion (40%) of Black first-generation students whose parents attended college but did not complete a degree.

>>>>

### Decentralized Diversity in a Multi-Nucleo Landscape

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles -- 11 Geomarkets, results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), results for which are shown in the [Online Appendix](https://ozanj.github.io/cb_geo/#/results). The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core (CITE- Harris and Ullman 1945). As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

The population growth in the Los Angeles metropolitan area increased from approximately 8 Million in 1980 to more than 12 Million in 2020. @fig-la-maps showcases population growth by Geomarkets over the three census periods reflect the urban sprawl forming the multi-nuclei model. The most pronounced increases occurred in South and South Central Los Angeles Geomarket, as well as the outer/inland Geomarkets of San Fernando Valley-West, Riverside and San Bernadino, and Ontario. The South and South Central Los Angeles Geomarket grew from approximately 1.2 to 1.6 Million from 1980 to 2020. In contrast, the San Fernando Valley-West and Riverside and San Bernadino, and Ontario Geomarkets, respectively, grew from approximately 700,000 to 1.2 Million and from 1.1 Million to nearly 3 Million during this time.

@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. Over this period, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. \[THE CAVEAT THAT WE DON'T HAVE ASIAN FOR 1980 IS MORE IMPORTANT FOR LA; KARINA -- MENTION THIS OR REVISE HOW YOU WRITE FINDINGS IN THIS PARAGRAPH, LIKE FOCUS MAINLY ON 2000-2020. YOU DECIDE.\] In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 62% White in Long Beach to 78% in the San Fernando Valleys (West and East) Geomarkets. By 2020, however, only the West Los Angeles and West Beach Geomarket remained predominantly White (62%). Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (44% White, 4% Black, 39% Hispanic; 11% Asian Pacific Islander), San Fernando Valley -East (43% White, 4% Black, 41% Hispanic; 8% Asian Pacific Islander), Glendale and Pasadena (35% White, 3% Black, 37% Hispanic; 21% Asian Pacific Islander), South Bay (29% White, 12% Black, 33% Hispanic; 21% Asian Pacific Islander), Long Beach (26% White, 11% Black, 46% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 3% Black, 52% Hispanic; 24% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 54% Hispanic; 6% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 18% and 18% to 9%, respectively.

@@

<!--# KARINA -- CUT THIS CUZ BASICALLY CONSISTENT WITH INCOME.
@fig-la-rq1-ses also illustrates parallel trends in poverty and educational attainment over time, with Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles Geomarkets consistenly ranking at lower levels across these socioeconomic indicators. Moreover, the disparities between these three Geomarkets and all others in the Los Angeles metropolitan area widened from 1980 to 2020.

 In 1980, poverty rates were relatively low (less than 12%) for all Geomarkets except for East Los Angeles (17%), South and South Central Los Angeles (24%), and Hollywood and Wilshire (18%). However, poverty rates declined in South and South Central Los Angeles (22%), remained stable in East Los Angeles and Hollywood and Wilshire, and increased in all other Geomarkets. Such shifts and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in more staggered poverty rates across the metropolitan area by 2020. This includes four Geomarkets with poverty rates less than 10% (San Fernando Valley- West, West Los Angeles and West Beach, South Bay, and Covina and West Covina), four Geomarkets with poverty rates ranging from 11%-15% (San Fernando Valley- East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario), and three Geomarkets with poverty rates ranging from 16%-22% (Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles). -->

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets in the Los Angeles metropolitan areas, we analyze six orders placed by a research university that filtered for prospects across the entire state of California, in the 2020 through 2022 high school graduating classes, and by PSAT scores. Two orders indicated PSAT score thresholds ranging from 1070-1180, another two orders filtered for a 1190-1260 PSAT range, and the remaining two orders filtered for scores ranging from 1270-1520.

@fig-la-rq2-race presents the racial/ethnic composition of purchased student profiles. For the more than 15,000 prospects in the resulting student lists at the low PSAT score thresholds (1070-1180), South and South Central Los Angeles (n=1,031) and East Los Angeles (n=871) Geomarket pools stand out as having an overwhelming majority of students of color, 98% and 95% respectively. Relative to other Geomarkets in the Los Angeles metropolitan area, such patterns are attributed to both South and South Central Los Angeles (89%) and East Los Angeles (64%) having substantially greater proportions of Hispanic prosepcts, as well as a larger proportion of Black prospects for South and South Central Los Angeles (6%) and larger share of Asian prospects for East Los Angeles (30%). Even among higher PSAT score ranges, South and South Central Los Angeles remains the most prominent Geomarket with significant proportions of Hispanic and Black student prospects, while East Los Angeles stands out for its substantial share of Asian prospects.

@@

@fig-la-rq2-midSAT-race shows how much each Geomarket accounts for the share of prospects in the middle PSAT score range by race/ethnicity. South and South Central Los Angeles accounts for only 2% of all purchased prospect profiles but accounts for 7% of Black prospects. The Reiverside, San Bernardino, \& Ontario Geomarket accounts for 17% of all prospects but 25% of Black prospects. For Hispanic prospects, South and South Central Los Angeles (8% versus 2% for all prospects) and Riverside, San Bernadino, and Ontario (24% versus 17% for all prospects) also contributed a disproportinately larger proportions of prospects relative to their contributions to the overall pool in the Los Angeles metropolitan area. For Asian students, the Glendale and Pasadena (17% of Asian prospects versus 14% for all prospects), East Los Angeles (11% of Asian prospects versus 5% of all prospects) and Covina and West Covina (17% of Asain prospects versus 11% of all prospects) Geomarkets contributed a disproportionate larger share of Asian prospects relative to their contributions to the overall prospect pool. For the more than 2,500 White prospects in this pool, the San Fernando Valley -- West (28% of White prospects versus 20% of all prospects) and the West Los Angeles and West Beach (12% of White prospects versus 7% for all prospects) Geomarkets contribute disproportionately larger shares.

These results demonstrate that following the @RN4982 strategy of filtering on high-income Geomarkets and excluding low-income Geomarkets -- such as the South and South Central Los Angeles Geomarket and the Riverside, San Bernadino, and Ontario Geomarket -- would result in the disproportionate exclusiong of Black and Hispanic students from the resulting prospect pools.Similarly, using a Geomarket filter excluding East Los Angeles would result in the loss of disproportionate shares of Asian students.

[KARINA -- CHECK YOUR NUMBERS FOR THIS PARAGRAPH] @fig-la-rq2-midSAT-ses presents Geomarket contributions by first-generation college student status, demonstrating the disproportionately larger share of first-generation college students contributed to the overall pool by East Los Angeles, South and South Central Los Angeles, and Riverside, San Bernadino, and Ontario Geomarkets. For students whose parents did not attend college, East Los Angeles (16% versus 5% of all prospects), South and South Central Los Angeles (10% versus 2% of all prospects), and Riverside, San Bernardino, and Ontario (24% versus 17% of all prospects) Geomarkets provided a disproportionately larger contribution of first-generation college students. <!--#  KARINA -- I CUT THIS TEXT. OK? These three Geomarkets also made disproportionately larger contributions of first-generation college student prospects whose parents attended but did not complete their degree, as well as Hollywood & Wilshire (7% versus 5% of all prospects), Long Beach (7% versus 5% of all prospects), and Covina and West Covina (13% versus 10% of all prospects).  -->

<!--#  [I THINK THE WAY YOU WRITE UP THIS FIGURE SHOULD BE CHANGED MORE SUBSTANTIALLY; I ADD SOME COMMENTS BELOW] --> Lastly, @fig-la-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Los Angeles metropolitan area. Across all Hispanic prospects in the middle range PSAT orders, 25% were first-generation college students whose parents did not attend college, 25% were first-generation college students whose parents attended some college but did not complete their degrees, and 50% were not first-generation college students. By contrast, for the 156 Hispanic prospects from the South and South Central Geomarket, 54% had parents who never attended college, 28% had parents who attended some college, and only 18% had parents with a BA. Similarly, Hispanic prospects from East Los Angeles, from Hollywood & Wilshire, and from Riverside, San Bernardino, and Ontario were much more likely to have parents without a BA compared to all Hispanic prospects in the Los Angeles metro area.

@@

While the pool of Black prospects is relatively small (n=186) in @fig-la-rq2-midSAT-combo, some patterns emerge by first-generation status and Geomarket. Overall, approximately 5% of Black prospects were first-generation college students-no college, 20% were first-generation college students-some college, and 74% not first-generation). The South and South Central Los Angeles Geomarket stands out for contributing a disproportionately high share (21%) of Black first-generation college students whose parents did not attend college, despite representing a relatively smaller overall number. Additionally, the Long Beach Geomarket is notable for its higher proportion (40%) of Black first-generation students whose parents attended college but did not complete a degree.
